Labor Market Income and Substitution Effects Explain Contemporary Natural Selection

### 2020-12-18 why-negative-selection.Rmd
Use polygenic scores not PGS

```{r plot-income-educ-level, fig.cap = "Selection effects by education and income. Each point represents a single bivariate regression of number of children on a PGS", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}

@@

Figure \@ref(fig:plot-income-educ-level) plots selection effects for
each polygenic score, grouping respondents by age of completing
full-time education, and by household income. Effects are larger and
more significant for the lowest income category, and for the lowest
education category. Note that the overall effect is not a simple average
of the effect among the different subgroups, because polygenic scores
may also shift respondents between the subgroups. For example, a high
PGS for educational attainment may predict having fewer children among
early school leavers, but it may also increase the age at which a
respondent leaves school.

@@

Selection effects are also different between men and women (Appendix
Figure \@ref(fig:plot-sexes)). Differences are particularly large for
educational attainment, height and MDD. Several PGS for mental illness
and personality traits are more selected for (or less against) among
women, including major depressive disorder (MDD), schizophrenia and
neuroticism, while extraversion is more selected for among men. PGS for waist
circumference and waist-hip ratio are less selected for among women, and PGS for
educational attainment are more selected against, though the difference
is only significant for the educational attainment PGS of @lee2018gene
("EA3"). One possible reason for these sex differences is
that polygenic scores may affect fertility via success in marriage markets, and
men and women may value different characteristics in these.

@@

Lastly, we split female respondents by age at first live birth
(AFLB).[^aflb-no-men] There is evidence for genetic effects on AFLB
[@Barban_2016], and there is a close link between this variable and number of
children born. Figure \@ref(fig:plot-age-flb)a shows effect sizes estimated
separately for each tercile of AFLB. Several effects are strikingly different
across terciles. ADHD and MDD are selected for amongst the youngest third of
mothers, but selected against among the oldest two-thirds. Educational
attainment is selected for among the oldest two-thirds of mothers, but is not
significantly selected among the youngest third. Similarly, several PGS of body
measurements are selected against only among older mothers. The correlation
between effect sizes for the youngest and oldest terciles is `r cor_age_flb`. As
before, recall that polygenic scores may have two effects: changing the number
of children within each category, and changing the number of children by
shifting respondents between categories.

@@

To investigate this further, we regress number of children on PGS
*controlling* for AFLB (Figure \@ref(fig:plot-age-flb)b). In
`r opp_signed` out of `r length(score_names)` cases, effects
change sign when controls are added. The correlation between effect
sizes controlling for AFLB, and raw effect sizes, is `r cor_raw_flb`.
Thus, selection effects seem to come through two opposing channels: an
effect on AFLB, and an opposite-signed effect on number of children 
controlling for AFLB.

@@

```{r plot-cors-earnings-educ, fig.cap="Selection effects, with and without controls for age at first live birth, by correlations with earnings and educational attainment. Each point represents one PGS.", fig.height = 4}

@@

We test this explanation in two ways. First, the economic theory predicts that
genetic variants will be selected for (or against) in proportion to their effect
on earnings. Figure \@ref(fig:plot-cors-earnings-educ) plots selection effects
on each PGS against that PGS's correlation with earnings in a respondent's first
job, and against its correlation with educational attainment, a
predictor of lifetime earnings. The raw relationships (left column) are strongly
negative. If we plot selection effects controlling for age at first live birth
(right column), the effect reverses and becomes positive. Thus, the labour
market appears to play an important role in natural selection. Substitution
effects dominate income effects overall, which fits the known association
between income and lower fertility [@becker1960economic;@jones2006economic].

@@

Second, we re-estimate equation \@ref(eq:regression) for each
polygenic score, controlling for earnings in first job and education levels. Effect sizes
are generally reduced (Appendix Figure \@ref(fig:plot-res-earn-educ)), and only
`r n_05_33` out of 33 PGS are significant at $p = 0.05/33$. This suggests that
earnings and education indeed mediate the effect of polygenic scores on fertility.

@@

An alternative theory is that traits selected for are linked to externalizing
behaviour, risk-seeking and low time discount rates, via the channel of early
sexual behaviour [@Mills2020.05.06.081273]. The data here provide some support
for this: scores which might plausibly be linked to externalizing behaviour,
like ADHD and age at smoking initiation, are selected for. However, this theory
is less good at explaining variation in selection across the full range of
scores, including physical measures, e.g. waist-hip ratio and BMI. Externalizing
behaviour also does not explain why selection effects should work in the
opposite direction among older parents, whereas the economic theory can explain
this via income effects, as described above. We test the alternative theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040). While risk attitude is always a
highly significant predictor of number of children, it has little impact on the
effect size or significance of the PGS. The median ratio of PGS effect sizes
between regressions with and without controls is `r median(ratio)`; all PGS
which are significant at $p \le 0.05/33$ in uncontrolled regressions remain so
when controlling for risk attitude. Thus, although risk attitude does predict
number of children, it appears not to mediate selection effects. Overall, we
believe that the economic theory is the most likely explanation.

>>>>

```{r plot-income-educ-level, fig.cap = "Selection effects by education and income. Each point represents a single bivariate regression of number of children on a polygenic score", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}

@@

Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
polygenic score, grouping respondents by age of completing full-time education,
and by household income. Effects are larger and more significant for the lowest
income category, and for the lowest education category. Note that the overall
effect is not a simple average of the effect among the different subgroups,
because polygenic scores may also shift respondents between the subgroups. For
example, a high polygenic score for educational attainment may predict having
fewer children among early school leavers, but it may also increase the age at
which a respondent leaves school.

@@

Selection effects are also different between men and women (Appendix Figure
\@ref(fig:plot-sexes)). Differences are particularly large for educational
attainment, height and MDD. Several polygenic scores for mental illness and
personality traits are more selected for (or less against) among women,
including major depressive disorder (MDD), schizophrenia and neuroticism, while
extraversion is more selected for among men. Scores for waist circumference and
waist-hip ratio are less selected for among women, and scores for educational
attainment are more selected against, though the difference is only significant
for the "EA3" educational attainment PGS of @lee2018gene . One possible reason
for these sex differences is that polygenic scores may affect fertility via
success in marriage markets, and men and women may value different
characteristics in these.

@@

Lastly, we split female respondents by age at first live birth
(AFLB).[^aflb-no-men] There is evidence for genetic effects on AFLB
[@Barban_2016], and there is a close link between this variable and number of
children born. Figure \@ref(fig:plot-age-flb)a shows effect sizes estimated
separately for each tercile of AFLB. Several effects are strikingly different
across terciles. ADHD and MDD are selected for amongst the youngest third of
mothers, but selected against among the oldest two-thirds. Educational
attainment is selected for among the oldest two-thirds of mothers, but is not
significantly selected among the youngest third. Similarly, several polygenic
scores for body measurements are selected against only among older mothers. The
correlation between effect sizes for the youngest and oldest terciles is `r
cor_age_flb`. As before, recall that polygenic scores may have two effects:
changing the number of children within each category, and changing the number of
children by shifting respondents between categories.

@@

To investigate this further, we regress *number of children* on polygenic scores
*controlling* for AFLB (Figure \@ref(fig:plot-age-flb)b). In `r opp_signed` out
of `r length(score_names)` cases, effects change sign when controls are added.
The correlation between effect sizes controlling for AFLB, and raw effect sizes,
is `r cor_raw_flb`. Thus, selection effects seem to come through two opposing
channels: an effect on AFLB, and an opposite-signed effect on number of children
controlling for AFLB.

@@

```{r plot-cors-earnings-educ, fig.cap="Selection effects, with and without controls for age at first live birth, by correlations with earnings and educational attainment. Each point represents one polygenic score.", fig.height = 4}

@@

We test this explanation in two ways. First, the economic theory predicts that
genetic variants will be selected for (or against) in proportion to their effect
on earnings. Figure \@ref(fig:plot-cors-earnings-educ) plots selection effects
on each polygenic score against that score's correlation with earnings in a respondent's first
job, and against its correlation with educational attainment, a
predictor of lifetime earnings. The raw relationships (left column) are strongly
negative. If we plot selection effects controlling for age at first live birth
(right column), the effect reverses and becomes positive. Thus, the labour
market appears to play an important role in natural selection. Substitution
effects dominate income effects overall, which fits the known association
between income and lower fertility [@becker1960economic;@jones2006economic].

@@

Second, we re-estimate equation \@ref(eq:regression) for each polygenic score,
controlling for earnings in first job and education levels. Effect sizes are
generally reduced (Appendix Figure \@ref(fig:plot-res-earn-educ)), and only 
`r n_05_33` out of 33 scores are significant at $p = 0.05/33$. This suggests that
earnings and education indeed mediate the effect of polygenic scores on
fertility.

@@

An alternative theory is that traits selected for are linked to externalizing
behaviour, risk-seeking and low time discount rates, via the channel of early
sexual behaviour [@Mills2020.05.06.081273]. The data here provide some support
for this: scores which might plausibly be linked to externalizing behaviour,
like ADHD and age at smoking initiation, are selected for. However, this theory
is less good at explaining variation in selection across the full range of
scores, including physical measures, e.g. waist-hip ratio and BMI. Externalizing
behaviour also does not explain why selection effects should work in the
opposite direction among older parents, whereas the economic theory can explain
this via income effects, as described above. We test the alternative theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040). While risk attitude is always a
highly significant predictor of number of children, it has little impact on the
effect size or significance of the polygenic scores. The median ratio of effect sizes
between regressions with and without controls is `r median(ratio)`; all scores
which are significant at $p \le 0.05/33$ in uncontrolled regressions remain so
when controlling for risk attitude. Thus, although risk attitude does predict
number of children, it appears not to mediate selection effects. Overall, we
believe that the economic theory is the most likely explanation.
